A Classical Density-Functional Theory for Describing Water Interfaces

### 2012-06-20 papers/water-SAFT/paper.tex

Statistical Associating Fluid Theory (SAFT) is a set of models for
complex fluids in which hydrogen bonding plays a significant
role\cite{muller2001molecular}.  SAFT is used to accurately model the
equations of state of pure fluids and predict the equations of state
for mixtures over a wide range of temperatures and pressures.  SAFT is
based on Wertheim's first-order thermodynamic perturbation theory
(TPT1) \cite{wertheim1984fluidsI, wertheim1984fluidsII,
  wertheim1986fluidsIII, wertheim1986fluidsIV} which allows it to
account for strong associative interactions between molecules.

The SAFT Helmholtz free energy is composed of five terms:
\begin{equation} \label{eq:SAFT-free-energy}
  F[n] = F_\textit{id} + F_\textit{hs} + F_\textit{disp} +
  F_\textit{assoc} + F_\textit{chain},
\end{equation}
where the first three terms encompass the \emph{monomer} contribution
to the free energy, the fourth is the \emph{association} free energy,
describing hydrogen bonds, and the final term is the chain formation
energy of for fluids that are chain polymers.  The final term is not
relevant for water.  While a number of SAFT formulations have been
published, we will focus on SAFT-VR\cite{gil-villegas-1997-SAFT-VR},
which has been used by Clark \emph{et al} to construct an optimal SAFT
model for water\cite{clark2006developing}. A majority of the empirical
parameters used in our functional are taken from the Clark \emph{et al}
paper which contains five empirical parameters and gives
correct pressures at the liquid-vapor coexistence curve shown in Figure
\ref{fig:pressure-with-isotherms}.

@@

SAFT has also been used to construct classical density functionals,
which have been applied to study the surface tension as a function of
temperature\cite{clark2006developing, gloor2004accurate,
  gloor2007prediction, blas2001examination, gloor2002saft}.  Such
functionals have qualitatively predicted the dependence of surface
tension on temperature, however, they also overestimate
the surface tension by a factor of about 1.5, and most are unsuited
for studying systems that have density variations on a molecular
length scale due to the use of a local density
approximation\cite{gloor2002saft,clark2006developing, gloor2007prediction,
gloor2004accurate, gross2009density, kahl2008modified, blas2001examination}.

@@

Functionals constructed using a local density approximation
will always fail to satisfy the contact-value theorem, and will
therefore incorrectly model small hard solutes.
The contact-value theorem states that the pressure of any fluid at a
hard wall interface is proportional to the contact density of the
fluid\cite{henderson1979exact}, according to the following relation:
\begin{equation}\label{eq:contact}
  p = n_ck_BT
\end{equation}
where $n_c$ is the contact density, $k_B$ is the Boltzmann constant
and $T$ is the temperature of
the fluid. This leads to the property that the free energy of
solvation of a small hard solute is proportional to the
solvent-excluded volume:
\begin{equation}\label{contactvaluethm}
  F = n k_BT V
\end{equation}
The contact-value theorem is satisfied for functionals where the
only purely local term is the free energy of an ideal gas.

\section{Theory and Methods}
We construct a new classical density functional for water. We treat
each individual water molecule as a purely classical hard sphere and
add association terms based on SAFT. For our free energy, we use the
first four terms from Equation \ref{eq:SAFT-free-energy}: $F_\textit{id}$,
$F_\textit{hs}$, $F_\textit{disp}$ and $F_\textit{assoc}$. In
addition, a chemical potential term is used, so we actually work in
the grand canonical ensemble.  In the following sections, we will
introduce the terms of this functional.

\subsection{Ideal gas functional}
The first term is the ideal gas free energy functional, which
incorporates effects of kinetic energy.  The ideal gas functional is
given by
\begin{equation}\label{idealgas}
  F_{id}[n] = k_B T \int n(\xx)\left( \ln{\frac{n(\xx)}{n_Q}} - 1\right) d\xx
\end{equation}
where n(\xx) is the density of water molecules and $n_Q$ is the
quantum concentration
\begin{equation}\label{quantumconcentration}
 n_Q =\left(\frac{mk_BT}{2\pi\hbar^2}\right)^{3/2}.
\end{equation}
The ideal gas free energy functional satisfies the contact value
theorem and its limiting case of small solutes (Equations
\ref{eq:contact} and \ref{contactvaluethm}). These properties are
retained by the total functional, provided the remaining terms are
\emph{purely nonlocal}.

@@

We treat the repulsive interactions using the White Bear version of
the Fundamental-Measure Theory~(FMT) functional for the hard-sphere
fluid~\cite{roth2002whitebear}.  In the homogeneous limit, the White
Bear functional reduces to the Mansoori-Carnahan-Starling-Leland
equation of state.  FMT functionals are expressed as the integral of
the \emph{fundamental measures} of a fluid, which provide local
measures of quantities such as the filling fraction, density of
spheres touching a given point and mean curvature.  The excess free
energy is written as:
\begin{equation}
F_{hs}[n] = k_B T \int (\Phi_1(\xx) + \Phi_2(\xx) + \Phi_3(\xx)) d\xx \; ,
\end{equation}
with integrands
\begin{align}
\Phi_1 &= -n_0 \ln\left( 1 - n_3\right)\\
\Phi_2 &= \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3} \\
\Phi_3 &= (n_2^3 - 3n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2})
  \frac{
    n_3 + (1-n_3)^2\ln(1-n_3)
  }{
    36\pi n_3^2\left( 1 - n_3 \right)^2
  } ,
\end{align}
where the fundamental measure densities are given by:
\begin{align}
  n_3(\xx) &= \int n(\xx') \Theta(\left|\xx - \xx'\right| - R) d\xx' \\
  n_2(\xx) &= \int n(\xx') \delta(\left|\xx - \xx'\right| - R) d\xx'
  \\
  \mathbf{n}_{V2} &= \mathbf{\nabla} n_3 \\
  n_1 &= \frac{n_2}{4\pi R}\\
  \mathbf{n}_{V1} &= \frac{\mathbf{n}_{V2}}{4\pi R}\\
  n_0 &= \frac{n_2}{4\pi R^2}.
\end{align}
The density $n_3$ is the filling fraction and $n_2$ describes the number
of spheres touching a given point. For our water functional, we use the 
hard-sphere radius of
3.03420~\AA, which was found to be optimal by Clark
\emph{et al}.\cite{clark2006developing}

\newcommand\etadisp{\ensuremath{\eta_\textit{d}}}
\newcommand\epsilondisp{\ensuremath{\epsilon_\textit{d}}}
\newcommand\epsilonassoc{\ensuremath{\epsilon_\textit{a}}}
\newcommand\kappaassoc{\ensuremath{\kappa_\textit{a}}}
\newcommand\lambdadisp{\ensuremath{\lambda_\textit{d}}}
\newcommand\lscale{\ensuremath{s_d}}
\subsection{Dispersion free energy}
The first term in the attractive energy is the dispersion term.  This
will include van der Waals attraction and any orientation-independent
interactions. We use a dispersion term based on the SAFT-VR
approach\cite{gil-villegas-1997-SAFT-VR}, which has two free
parameters: an interaction energy $\epsilondisp$ and a
length scale $\lambdadisp R$.

The SAFT-VR dispersion free energy has the form~\cite{gil-villegas-1997-SAFT-VR}
\begin{align}
  F_\text{disp}[n] &= \int \left(a_1(\xx) + \beta a_2(\xx)\right)n(\xx)d\xx
\end{align}
where $a_1$ and $a_2$ are the first two terms in a high-temperature
perturbation expansion and $\beta=1/k_BT$.  The first term, $a_1$, is 
the mean-field dispersion interaction and the second term, $a_2$, describes the
effect of fluctuations resulting from compression of the fluid due
to the dispersion interaction itself. $a_2$ is approximated
using the local compressibility approximation (LCA), which
assumes the energy fluctuation is directly related to the
compressibility of a hard-sphere reference fluid\cite{barker1976liquid}.

The form of $a_1$ and $a_2$ for SAFT-VR is given in
reference~\cite{gil-villegas-1997-SAFT-VR} but is expressed in terms
of the filling fraction.  In order to apply this form to an
inhomogeneous density distribution, we construct an effective local
filling fraction for dispersion $\etadisp$, given by
\begin{align}
%    ((4/3.0*M_PI)*Pow(3)(R)*GaussianConvolve(2*lambdainput*lscale*radius,
%                                             2*lambdaE*length_scalingE*Rexpr)
%  \etadisp(\xx) &= \color{red}
%  \int \frac{\Theta(2\lambdadisp\lscale R - \left|\xx-\xx'\right|)}{
%             \left(2\lambdadisp\lscale\right)^3} n(\xx') d\xx'.
  \\
  \etadisp(\xx) &= \frac{1}{6\sqrt{\pi} \lambdadisp^3\lscale^3}
  \int n(\xx')\exp\left({-\frac{|\xx-\xx'|^2}{2(2 \lambdadisp
      \lscale R)^2}}\right)d\xx'
\end{align}
This effective filling fraction is used throughout the dispersion
calculations, and represents a filling fraction averaged over the
effective range of the dispersive interaction.  This functional
introduces an additional empirical parameter $\lscale$ which adjusts
the length scale over which the dispersion interaction is correlated.

\subsection{Association free energy}
The final attractive energy term is the association term, which
accounts for hydrogen bonding.  Hydrogen bonds are modeled as four
attractive patches (``association sites'') on the surface of the hard
sphere.  These four sites represent two protons and two electron lone
pairs.  There is an attractive energy $\epsilonassoc$ when
two molecules are oriented such that the proton of one overlaps
with the lone pair of the other.  The volume over which this
interaction occurs is $\kappaassoc$, giving the association
term in the free energy two empirical parameters that are fit to
experimental data.

The association functional we use is a modified version of Yu
and Wu\cite{yu2002fmt-dft-inhomogeneous-associating}. \footnote{We
  should note that Fu and Wu\cite{fu2005vapor-liquid-dft} use almost
  the same functional, but their paper contains errors in the
  association term and is not useful as a reference.}
This functional includes the effects of density inhomogeneities in the
contact value of the correlation function $g^{HS}_\sigma$, but is
based on the SAFT-HS equation of state rather than the SAFT-VR
equation of state\cite{gil-villegas-1997-SAFT-VR}. SAFT-VR is used in the
optimal SAFT parametrization for water found by Clark \emph{et
  al}\cite{clark2006developing}, but to use it's
equation of state involves adding a correction term (see
Equation~\ref{eq:gSW}).

The association functional we use is constructed by using the density
$n_0(\xx)$, which is the density of hard spheres touching a given
point, in the standard SAFT-VR association
energy\cite{gil-villegas-1997-SAFT-VR}.
The association free energy for our four-site model has the form
\begin{align}
  F_\text{assoc}[n] &= 4 k_BT \int n_0(\xx) \zeta(\xx)
  \left(\ln X(\xx) - \frac{X(\xx)}{2} + \frac12\right) d\xx
\end{align}
where the value of $4$ comes from the four association sites per
molecule, and the functional $X$ is the fraction of association sites
\emph{not} hydrogen-bonded, and $\zeta(\xx)$ is a dimensionless
measure of the density inhomogeneity.
%
The fraction $X$ is determined by the quadratic equation
\begin{align}
  X(\xx) &= \frac{\sqrt{1 + 8n_0(\xx)\zeta(\xx)
      \Delta(\xx)} - 1}
  {4 n_0(\xx)\zeta(\xx)
    \Delta(\xx)}
\end{align}
where the factor $\zeta$ from Yu and
Wu\cite{yu2002fmt-dft-inhomogeneous-associating} describes the
inhomogeneity of the fluid density, and reduces to unity in the
homogeneous limit:
\begin{align}
  \zeta(\xx) &= 1 - \frac{\mathbf{n}_{2V}\cdot\mathbf{n}_{2V}}{n_2^2}.
\end{align}
The functional $\Delta$ which appears in the formula for $X$ is a
measure of hydrogen-bonding probability.  In the SAFT-VR
model\cite{gil-villegas-1997-SAFT-VR}, $\Delta$ is given by
\begin{align}
  \Delta(\xx) &= \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right) \\
  g^\textit{SW}_\sigma(\xx) &= g^\textit{HS}_\sigma\left(\frac{4\pi R^3}{3}n_0(\xx)\right) +
  \frac{1}{4}\beta\left(\frac{\partial a_1}{\partial \etadisp(\xx)} -
  \frac{\lambdadisp}{3 \etadisp}\frac{\partial a_1}{\partial \lambdadisp}\right)\label{eq:gSW}
\end{align}
% This is equation 77 of gil-villegas-1997-SAFT-VR
where $g^\textit{SW}_\sigma$ is the correlation function evaluated at
contact for a hard-sphere fluid with a square-well attractive
potential, and $a_1$ and $a_2$ are two terms in the dispersive term of
the free energy.  The correlation function $g^\textit{SW}_\sigma$ is
written as a perturbative correction to the hard-sphere fluid correlation
function $g^\textit{HS}_\sigma$, for which we use the functional of Yu and
Wu\cite{yu2002fmt-dft-inhomogeneous-associating}:
\begin{align}
  g_\sigma^{HS}(\xx) &= \frac{1}{1-\etadisp(\xx)}
  +\frac32\frac{\zeta(\xx)}{(1-\etadisp(\xx))^2}
  + \frac12\frac{\etadisp(\xx)^2\zeta(\xx)}{(1-\etadisp(\xx))^3}.
\end{align}

>>>>

Statistical Associating Fluid Theory (SAFT) is a theory describing
complex fluids in which hydrogen bonding plays a significant
role\cite{muller2001molecular}.  SAFT is used to accurately model the
equations of state of both pure fluids and mixtures over a wide range
of temperatures and pressures.  SAFT is
based on Wertheim's first-order thermodynamic perturbation theory
(TPT1)\cite{wertheim1984fluidsI, wertheim1984fluidsII,
  wertheim1986fluidsIII, wertheim1986fluidsIV}, which allows it to
account for strong associative interactions between molecules.

The SAFT Helmholtz free energy is composed of five terms:
\begin{equation} \label{eq:SAFT-free-energy}
  F[n] = F_\textit{id} + F_\textit{hs} + F_\textit{disp} +
  F_\textit{assoc} + F_\textit{chain},
\end{equation}
where the first three terms encompass the \emph{monomer} contribution
to the free energy, the fourth is the \emph{association} free energy,
describing hydrogen bonds, and the final term is the chain formation
energy of for fluids that (unlike water) are chain polymers.
While a number of forumlations of SAFT have been
published, we will focus on SAFT-VR\cite{gil-villegas-1997-SAFT-VR},
which was used by Clark \emph{et al} to construct an optimal SAFT
model for water\cite{clark2006developing}. All but one of the six
empirical parameters used in the functional introduced in this paper
are taken directly from this Clark \emph{et al} paper.  The phase
diagram for this optimal SAFT model for water is shown in
Figure~\ref{fig:pressure-with-isotherms}, which demonstrates that its
vapor pressure as a function of temperature is very accurate, while
the liquid density shows larger discrepancies.  The critical point is
very poorly described, which is a common failing of mean-field
expansions.

@@

SAFT has also been used to construct classical density functionals,
which are commonly used to study the surface tension as a function of
temperature\cite{clark2006developing, gloor2004accurate,
  gloor2007prediction, blas2001examination, gloor2002saft}.  Such
functionals have qualitatively predicted the dependence of surface
tension on temperature, but they also overestimate
the surface tension by about 50\%, and most SAFT-based functionals are
unsuited for studying systems that have density variations on a
molecular length scale due to the use of a local density
approximation\cite{gloor2002saft,clark2006developing, gloor2007prediction,
gloor2004accurate, gross2009density, kahl2008modified, blas2001examination}.

@@

Functionals constructed using a local density approximation
fail to satisfy the contact-value theorem, and
therefore incorrectly model small hard solutes.
The contact-value theorem states that the pressure of any fluid at a
hard wall interface is proportional to the contact density of the
fluid\cite{henderson1979exact}:
\begin{equation}\label{eq:contact}
  p = n_ck_BT
\end{equation}
where $n_c$ is the contact density, $k_B$ is the Boltzmann constant
and $T$ is the temperature of
the fluid. This leads to the property that the free energy of
solvation of a small hard solute is proportional to the
solvent-excluded volume:
\begin{equation}\label{contactvaluethm}
  F = n k_BT V
\end{equation}
The contact-value theorem is satisfied for functionals where the
only purely local term is the free energy of an ideal gas, and is not
satisfied for functionals built using a local density approximation.

\section{Theory and Methods}
We construct a classical density functional for water, which reduces
in the homogeneous limit to the optimal SAFT model for water found by
Clark~\emph{et al}.  The Helmholtz free energy is constructed using the
first four terms from Equation \ref{eq:SAFT-free-energy}: $F_\textit{id}$,
$F_\textit{hs}$, $F_\textit{disp}$ and $F_\textit{assoc}$.  In the
following sections, we will introduce the terms of this functional.

\subsection{Ideal gas functional}
The first term is the ideal gas free energy functional, which is
purely local:
\begin{equation}\label{idealgas}
  F_{id}[n] = k_B T \int n(\xx)\left( \ln{\frac{n(\xx)}{n_Q}} - 1\right) d\xx
\end{equation}
where n(\xx) is the density of water molecules and $n_Q$ is the
quantum concentration
\begin{equation}\label{quantumconcentration}
 n_Q =\left(\frac{mk_BT}{2\pi\hbar^2}\right)^{3/2}.
\end{equation}
The ideal gas free energy functional on its own satisfies the contact
value theorem and its limiting case of small solutes (Equations
\ref{eq:contact} and \ref{contactvaluethm}). These properties are
retained by the total functional, provided all the remaining terms are
\emph{purely nonlocal}.

@@

We treat the hard-sphere repulsive interactions using the White Bear version of
the Fundamental-Measure Theory~(FMT) functional for the hard-sphere
fluid~\cite{roth2002whitebear}.  In the homogeneous limit, the White
Bear functional reduces to the Carnahan-Starling
equation of state.  FMT functionals are expressed as the integral of
the \emph{fundamental measures} of a fluid, which provide local
measures of quantities such as the filling fraction, density of
spheres touching a given point and mean curvature.  The hard-sphere
excess free energy is written as:
\begin{equation}
F_{hs}[n] = k_B T \int (\Phi_1(\xx) + \Phi_2(\xx) + \Phi_3(\xx)) d\xx \; ,
\end{equation}
with integrands
\begin{align}
\Phi_1 &= -n_0 \ln\left( 1 - n_3\right)\\
\Phi_2 &= \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3} \\
\Phi_3 &= (n_2^3 - 3n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2})
  \frac{
    n_3 + (1-n_3)^2\ln(1-n_3)
  }{
    36\pi n_3^2\left( 1 - n_3 \right)^2
  } ,
\end{align}
where the fundamental measure densities are given by:
\begin{align}
  n_3(\xx) &= \int n(\xx') \Theta(\left|\xx - \xx'\right| - R) d\xx' \\
  n_2(\xx) &= \int n(\xx') \delta(\left|\xx - \xx'\right| - R) d\xx'
  \\
  \mathbf{n}_{V2} &= \mathbf{\nabla} n_3 \\
  n_1 &= \frac{n_2}{4\pi R}\\
  \mathbf{n}_{V1} &= \frac{\mathbf{n}_{V2}}{4\pi R}\\
  n_0 &= \frac{n_2}{4\pi R^2}.
\end{align}
The density $n_3$ is the filling fraction and $n_2$ describes the number
of spheres touching a given point. For our water functional, we use the 
hard-sphere radius of
3.03420~\AA, which was found to be optimal by Clark
\emph{et al}.\cite{clark2006developing}

\newcommand\etadisp{\ensuremath{\eta_\textit{d}}}
\newcommand\epsilondisp{\ensuremath{\epsilon_\textit{d}}}
\newcommand\epsilonassoc{\ensuremath{\epsilon_\textit{a}}}
\newcommand\kappaassoc{\ensuremath{\kappa_\textit{a}}}
\newcommand\lambdadisp{\ensuremath{\lambda_\textit{d}}}
\newcommand\lscale{\ensuremath{s_d}}
\subsection{Dispersion free energy}
The dispersion free energy includes the van der Waals attraction and
any orientation-independent interactions. We use a dispersion term
based on the SAFT-VR approach\cite{gil-villegas-1997-SAFT-VR}, which
has two free parameters (taken from Clark~\emph{et
  al}\cite{clark2006developing}): an interaction energy $\epsilondisp$
and a length scale $\lambdadisp R$.

The SAFT-VR dispersion free energy has the form~\cite{gil-villegas-1997-SAFT-VR}
\begin{align}
  F_\text{disp}[n] &= \int \left(a_1(\xx) + \beta a_2(\xx)\right)n(\xx)d\xx
\end{align}
where $a_1$ and $a_2$ are the first two terms in a high-temperature
perturbation expansion and $\beta=1/k_BT$.  The first term, $a_1$, is 
the mean-field dispersion interaction and the second term, $a_2$, describes the
effect of fluctuations resulting from compression of the fluid due
to the dispersion interaction itself. $a_2$ is approximated
using the local compressibility approximation (LCA), which
assumes the energy fluctuation is simply related to the
compressibility of a hard-sphere reference fluid\cite{barker1976liquid}.

The form of $a_1$ and $a_2$ for SAFT-VR is given in
reference~\cite{gil-villegas-1997-SAFT-VR}, expressed in terms
of the filling fraction.  In order to apply this form to an
\emph{inhomogeneous} density distribution, we construct an effective local
filling fraction for dispersion $\etadisp$, given by a Gaussian
convolution of the density:
\begin{align}
%    ((4/3.0*M_PI)*Pow(3)(R)*GaussianConvolve(2*lambdainput*lscale*radius,
%                                             2*lambdaE*length_scalingE*Rexpr)
%  \etadisp(\xx) &= \color{red}
%  \int \frac{\Theta(2\lambdadisp\lscale R - \left|\xx-\xx'\right|)}{
%             \left(2\lambdadisp\lscale\right)^3} n(\xx') d\xx'.
  \etadisp(\xx) &= \frac{1}{6\sqrt{\pi} \lambdadisp^3\lscale^3}
  \int n(\xx')\exp\left({-\frac{|\xx-\xx'|^2}{2(2 \lambdadisp
      \lscale R)^2}}\right)d\xx'
\end{align}
This effective filling fraction is used throughout the dispersion
functional, and represents a filling fraction averaged over the
effective range of the dispersive interaction.  This functional
introduces an additional empirical parameter $\lscale$ which adjusts
the length scale over which the dispersion interaction is correlated.

\subsection{Association free energy}
The final attractive energy term is the association term, which
accounts for hydrogen bonding.  Hydrogen bonds are modeled as four
attractive patches (``association sites'') on the surface of the hard
sphere.  These four sites represent two protons and two electron lone
pairs.  There is an attractive energy $\epsilonassoc$ when
two molecules are oriented such that the proton of one overlaps
with the lone pair of the other.  The volume over which this
interaction occurs is $\kappaassoc$, giving the association
term in the free energy two empirical parameters that are fit to
experimental data (again, taken from Clark~\emph{et
  al}\cite{clark2006developing}).

The association functional we use is a modified version of Yu
and Wu\cite{yu2002fmt-dft-inhomogeneous-associating}, which
% \footnote{We
%  should note that Fu and Wu\cite{fu2005vapor-liquid-dft} use almost
%  the same functional, but their paper contains errors in the
%  association term and is not useful as a reference.}
includes the effects of density inhomogeneities in the contact value
of the correlation function $g^{HS}_\sigma$, but is based on the
SAFT-HS model, rather than the SAFT-VR
model\cite{gil-villegas-1997-SAFT-VR}, which is used in the optimal
SAFT parametrization for water of Clark \emph{et
  al}\cite{clark2006developing}.  Adapting Yu and Wu's association
free energy to SAFT-VR, simply involves the addition of a correction
term in the correlation function (see Equation~\ref{eq:gSW}).

The association functional we use is constructed by using the density
$n_0(\xx)$, which is the density of hard spheres touching a given
point, in the standard SAFT-VR association
energy\cite{gil-villegas-1997-SAFT-VR}.
The association free energy for our four-site model has the form
\begin{align}
  F_\text{assoc}[n] &= 4 k_BT \int n_0(\xx) \zeta(\xx)
  \left(\ln X(\xx) - \frac{X(\xx)}{2} + \frac12\right) d\xx
\end{align}
where the value of $4$ comes from the four association sites per
molecule, and the functional $X$ is the fraction of association sites
\emph{not} hydrogen-bonded, and $\zeta(\xx)$ is a dimensionless
measure of the density inhomogeneity.
\begin{align}
  \zeta(\xx) &= 1 - \frac{\mathbf{n}_{2V}\cdot\mathbf{n}_{2V}}{n_2^2}.
\end{align}
%
The fraction $X$ is determined by the quadratic equation
\begin{align}
  X(\xx) &= \frac{\sqrt{1 + 8n_0(\xx)\zeta(\xx)
      \Delta(\xx)} - 1}
  {4 n_0(\xx)\zeta(\xx)
    \Delta(\xx)}
\end{align}
where the functional $\Delta$ which is a measure of hydrogen-bonding
probability, given by
\begin{align}
  \Delta(\xx) &= \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right) \\
  g^\textit{SW}_\sigma(\xx) &= g^\textit{HS}_\sigma(\xx) +
  \frac{1}{4}\beta\left(\frac{\partial a_1}{\partial \etadisp(\xx)} -
  \frac{\lambdadisp}{3 \etadisp}\frac{\partial a_1}{\partial \lambdadisp}\right)\label{eq:gSW}
\end{align}
% This is equation 77 of gil-villegas-1997-SAFT-VR
where $g^\textit{SW}_\sigma$ is the correlation function evaluated at
contact for a hard-sphere fluid with a square-well attractive
potential, and $a_1$ and $a_2$ are the two terms in the dispersion
free energy.  The correlation function $g^\textit{SW}_\sigma$ is
written as a perturbative correction to the hard-sphere fluid
correlation function $g^\textit{HS}_\sigma$, for which we use the
functional of Yu and Wu\cite{yu2002fmt-dft-inhomogeneous-associating}:
\begin{align}
  g_\sigma^{HS} &= \frac{1}{1-n_3}
  +\frac32\frac{\zeta\zeta_2}{(1-n_3)^2}
  + \frac12\frac{\zeta_2^2\zeta}{(1-n_3)^3} \\
  \zeta_2 &= \frac{R}{3}n_2
\end{align}
